Stage-Splitting a Modal Language

### 2014-09-05 paper/semantics.tex
Incorporated Kayvon's suggestions.

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:%
%\footnote{We render $\fut$ as {\tt \$} in code examples.}

@@

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block bound to it. 
Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->1+2]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[y|->1+2]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the remaining $\next$, we must first partially evaluate the body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->1+2]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->1+2]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = 1+2 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example, we represented this environment with
explicit substitutions. For all of the normal features of \lang
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on values as standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.

When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  We call this search process
\emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes
a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
speculated into a residual ($q$), we then return a fresh variable wrapped in a
$\next$ block ($\next~\hat y$), along with a future environment which maps that
variable to our residual ($\hat y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
nothing but recursively speculate into every subexpression.  Once speculation
finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which
produces a future context and (by canonical forms) a $\next$-wrapped variable
($\next\{\hat y\}$)\footnote{
Importantly, speculation will search into both branches of a stage-\bbtwo\ {\tt if}
statement to evaluate all contained stage-\bbone\ code.  Thus the name.
}.  The context is then reified (using the $\reifysym$
judgment) into a series of let bindings, with $\mathtt{\hat y}$---stripped of
its $\next$---at the bottom.

@@

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value, where $q$ is a residual,
then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

@@

The dynamics described in the previous section provide a formal description of partial evaluation.
Recall that in partial evaluation systems, we have a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."

>>>>

% Ignoring for the moment how to define recursive functions in \lang,
\noindent
The stage-\bbone\ term below defines the function {\tt fexp} that implements the fast exponentiation algorithm:

@@

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->1+2]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[y|->1+2]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->1+2]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->1+2]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = 1+2 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example above, we represented this environment with
explicit substitutions. For all of the normal features of \lang\ 
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on (partial) values as does standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.

When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  This search process, called
\emph{speculation}, is implemented by the $\redtwosym$ judgment, which takes
a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
speculated into a residual ($q$), the output of $\redonesym$ is a fresh variable wrapped in a
$\next$ block ($\next~\hat y$), along with a future environment which maps that
variable to the residual ($\hat y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
nothing but recursively speculate into every subexpression.  Once speculation
finds a $\prev$ block, it resumes stage \bbone\ evaluation of the contents, which
produces a future context and (by canonical forms) a $\next$-wrapped variable
($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
judgment) into a series of let bindings with $\mathtt{\hat y}$, stripped of
its $\next$, at the bottom.

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage-\bbone\ code. 
Thus the evaluation of that stage one code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."

@@

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value for any residual $q$,
then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

@@

The dynamics described in the previous section provide a formal description of partial evaluation.
Recall that in partial evaluation systems, we start with a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."

### 2014-09-05 paper/splitting.tex
Incorporated Kayvon's suggestions.

\subsection{ Expression Splitting}

@@

The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono\ has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.

@@

@@

We give the following end-to-end correctness lemmas for open terms. 
In some of the theorems, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.
Substitution splitting works just like value splitting, which is why they use the same symbol.

>>>>

\subsection{Expression Splitting}

@@

The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono\ has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.

@@

\subsubsection {Expression Splitting at Values}

Note that general expression splitting always produces an effectively trivial precomputation when applied to partial values,
where effectively trivial means that the precomputation is either a unit value, or a tuple or injection of an effectively trivial values.

Partial value splitting is more than just a restriction of general expression splitting 
precisely because it observes this property it at the structural level, doing away with the precomputation entirely.
This makes sense intuitively, since a precomputation would imply that there is some computation left to do, 
which is antithetical to the notion of values.

@@

We give the following end-to-end correctness lemmas for open terms. 
Those readers not interested in the gritty details are advised to skip ahead to the theorems for closed terms, which give the important gist.

In some of the lemmas, we sort the usually heterogeneous context into $\Gamma$ and $\Gamma'$, 
which respectively hold all of the stage \bbone\ and stage \bbtwo\ variable bindings.
Substitution splitting works just like value splitting, which is why they use the same symbol.
